---
title: "Bend Genetics Data Wrangling"
author: "Inigo Peng"
date: "2022-10-12"
output: html_document
---
# Bend Genetics Data
# Description of Data

**Collection method:**

Hydrolab DS5 Multiprobe, Sampling Bottles

**Raw File format:**

Bend Genetics CSV lab report

**Monitoring (Project ID)**

Clear Lake Harmful Algal Bloom (HAB), Tap Water Sampling (CalWATCH).

**WQX submittal required?**

Clear Lake Harmful Algal Bloom YES. Tap Water Sampling NO.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(janitor)
library(readxl)
library(stringr)
library(lubridate)
library(readr)
```

## Data transformations

Bend Genetic lab report results come in two data frames (Analytical Report For Samples and Sample Results). We need to parse them from the excel file. We will first read in the excel file, skipping the first 5 lines of the header and delete an empty column (x8). We have also changed the column type for Sample ID into numeric - any row without numbers will now be NA. This is import for parsing in the next step.

```{r}
bend_raw <- read_excel("data-raw/bend/BendHAB.xlsx", skip = 5) %>% 
  clean_names() %>% 
  mutate(sample_id = as.numeric(sample_id)) %>% 
  select(-x8) %>% 
  glimpse()
```

To parse the Analytical Report For Samples dataframe, we know that the data frame ends after "Sample ID 1839". The next row is NA. So we parse a dataframe where it ends before the row that is NA.

```{r}
analytical_report_for_samples <- bend_raw[1:which(is.na(bend_raw$sample_id))[1] - 1, ] %>% 
  glimpse
```

To parse for the Sample Results data frame, we know that it starts after the second NA row (excel row 28 "Sample ID"), and ends at the third NA (excel row 54 "Quality Control"). We then create the corresponding column names for the sample results dataframe.

```{r}
sample_results <- bend_raw[(which(is.na(bend_raw$sample_id))[2]) : (which(is.na(bend_raw$sample_id))[3] - 1), ] %>%
  row_to_names(row_number = 1) %>%
  clean_names() %>% 
  rename(sample_id = na) %>% glimpse
```

We then create a the full data frame by joining Analytical Report for Samples and Sample Results.
```{r}
# Seems like wqx only takes units in ug/L and not copies/mL. Method ELISA have resutls in ug/L.
bend_full_df <-   left_join(analytical_report_for_samples, sample_results) %>% 
  filter(method == "ELISA") %>% glimpse()
```

To prepare data transformation to WQX format, we create the make_activity_id() function.
```{r}
make_activity_id <- function(location_id, date, activity_type, equipment_name, depth = NULL, time = NULL) {
  YYYYMMDD <- gsub('/', '', date)
  activity <- ifelse(activity_type == "Sample-Routine", "SR", "FM")
  equipment <- case_when(
    equipment_name == "Probe/Sensor" ~ "PS",
    equipment_name == "Water Bottle" ~ "WB",
    TRUE ~ NA_character_)
  hhmm <- gsub(':', '', time)
  equipment_comment <- case_when(
    equipment_name == "Hydrolab Surveyor DS5 Multiprobe" ~ "Hydro",
    equipment_name == "AlgaeChek Ultra Fluorometer" ~ "Algae", 
    TRUE ~ "")
  depth <- ifelse(is.na(depth), "", depth)
  paste(location_id, YYYYMMDD, hhmm,activity, equipment, depth, equipment_comment, sep = ":")
}
```

Lastly we map the full data frame to WQX schema by creating rows based on the values of the bend genetics data frame.
```{r}
bend_wqx <- bend_full_df %>% 
  mutate("Project ID" = "HAB",
         # Not all locations are recorded in master monitoring plan:
        # Benthic LUC01, OA04, LA03
         "Monitoring Location ID" = location,
         "Activity ID User Supplied (PARENTs)" = NA,
         "Activity Type" = "Sample-Routine", 
         "Activity Media Name" = matrix,
         "Activity Start Date" = format(mdy_hm(date_collected), "%m/%d/%Y"),
         "Activity Start Time" = format(mdy_hm(date_collected), "%H:%M"),
         "Activity Start Time Zone" = "PST",
        # Need activity depth/height, unit
         "Activity Depth/Height Measure" = NA,
         "Activity Depth/Height Unit" = NA,
        # Confirm Sample Collection method id is BVR SWQAPP
         "Sample Collection Method ID" = "BVR SWQAPP",
         "Sample Collection Method Context" = "CA_BVR",
        # Confirm Equipment for bend is Water Bottle
         "Sample Collection Equipment Name" = "Water Bottle",
         "Sample Collection Equipment Comment" = NA,
        #Should we use microcystin/nod. or just microcystin
         "Characteristic Name" = target,
         "Characteristic Name User Supplied" = NA,
         "Method Speciation" = NA,
         "Result Detection Condition" = ifelse(result == "ND", "Not Detected", NA),
         "Result Value" = ifelse(result == "ND", NA, result),
        # Do we use copies/mL? do not see it in previous wqx upload
         "Result Unit" = ifelse(result == "ND", NA, units),
         "Result Measure Qualifier" = NA,
         "Result Sample Fraction" = "Total",
         "Result Status ID" = "Final",
         "ResultTemperatureBasis" = NA,
         "Statistical Base Code" = NA,
         "ResultTimeBasis" = NA,
         "Result Value Type" = "Actual",
        # Seems like we are only using ELSA analysis not QPCR?
         "Result Analytical Method ID" = "520011",
         "Result Analytical Method Context" = "ABRAXIS LLC",
         "Analysis Start Date" = format(mdy_hm(date_received), "%m/%d/%Y"),
         "Result Detection/Quantitation Limit Type" = "Practical Quantitation Limit",
         "Result Detection/Quantitation Limit Measure" = quantitation_limit,
         "Result Detection/Quantitation Limit Unit" = units,
         "Result Comment" = notes,
         "Activity ID (CHILD-subset)" = make_activity_id(location_id = location,
                                                        date = `Activity Start Date`,
                                                        time = `Activity Start Time`,
                                                        activity_type = `Activity Type`,
                                                        equipment_name = `Sample Collection Equipment Name`,
                                                        depth = `Activity Depth/Height Measure`)
  ) %>% 
  relocate("Activity ID (CHILD-subset)", .before = "Activity ID User Supplied (PARENTs)") %>% 
  select(-c(0:13)) %>% glimpse
```

And we export the data frame to a csv file using the function write.csv(). Note we need to change the file encoding to make sure some of the special characters are exported correctly.

```{r}
write.csv(bend_wqx, "data/bend_wqx.csv", na = "", fileEncoding="Windows-1252", row.names = FALSE)

```




