---
title: "Alpha Lab Data Cleaning and Upload"
runtime: shiny
output: 
  html_document:
    code_folding: hide
    
---
## Data Description



| | |
|:-|:-|
|**Colection Method**        |    Sampling Bottles|
|**Raw File Format**         | AlphaLab CSV lab report |
|**Monitoring (Project ID)** | Storm Water Sampling (SW); Tap Water Sampling (CalWATCH).|
|**WQX submittal required?** | Yes |
|**Rate of Upload** | None|



**Note:**

This script is a dynamic markdown where the user can interact with the document. To run the document, click on "Run Document" to render it in HTML format. 

Usually, to run a markdown script, the user must ensure the project is set to the appropriate working directory. However, since this is an interactive markdown, we do not have to worry about that. The only files we need are the Alpha Lab excel files that need to get updated.

In these markdown documents, click on the "code" button on the right to see the code used to process the data. If you have questions about any of the functions used, type in "? " followed by the function (eg. `?mutate`) in the console,  or google search the function to see detailed documentation.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      tidy.opts = list(width.cutoff = 60), tidy = TRUE)

library(tidyverse)
library(readxl) 
library(lubridate)
library(wqxWeb)
library(DT)
library(shiny)
library(stringr)
library(emo)
reticulate::use_condaenv("r-reticulate")

```
```{r, include=FALSE}
project_id_lookup <- c(
  "M1" = "MS",
  "M2" = "MS",
  "M3" = "MS",
  "M4" = "MS",
  "HSP" = "SW",
  "BVSWD1" = "SW",
  "NBPRSC" = "SW",
  "RSTCC" = "SW",
  "BVCL1" = "SW",
  "BVRTC1" = "SW",
  "BVRTCC" = "SW",
  "BVSWDRV" = "SW",#not in cdx
  "RVSI1" = "SW",#not in cdx
  "RVSI2" = "SW",#not in cdx
  "BVCL2" = "CLM",	
  "BVCL3" = "CLM",	
  "BVCL5" = "CLM",	
  "BVCL6" = "CLM",	
  "BVCL11" = "CLM",	
  "BVCL12" = "CLM",	
  "BVCL13" = "CLM",	
  "BVCL14" = "CLM",	
  "BVCL15" = "CLM",	
  "BVCL16" = "CLM",	
  "BVCL17" = "CLM",	
  "BVCL18" = "CLM",	
  "BVCL19" = "CLM",	
  "BVCL20" = "CLM",
  "FC1" = "CS", 
  "FC2" = "CS",
  "FC3" = "CS",#not in cdx
  "MC1" = "CS",
  "MC2" = "CS",
  "TC1" = "CS",
  "AC1" = "CS",
  "AC2" = "CS",
  "AC3" = "CS",
  "AC4" = "CS",
  "MCC1" = "CS",
  "KC1" = "CS",
  "CC1" = "CS",
  "SC1" = "CS",
  "SC2" = "CS",#not in cdx
  "SHC2"= "CS",#not in cdx
  "CC2" = "CS",#not in cdx
  "SIEG01" = "CS",#not in cdx
  "COOP01" = "CS",#not in cdx
  "CLOV01" = "CS",#not in cdx
  "DRY01" = "CS",#not in cdx
  "COY01" = "CS",#not in cdx
  "SCOT01"= "CS",#not in cdx
  "AND01" = "CS",#not in cdx
  "NFORK01" = "CS",#not in cdx
  "LONG01" = "CS", #not in cdx
  "PUT01" = "CS", #not in cdx
  "MID01"= "CS", #not in cdx
  "KC5"= "CS", #not in cdx
  "AP01" = "HAB",
  "BP" = "HAB",
  "CLOAKS01" = "HAB",
  "CLV7" = "HAB",
  "CP" = "HAB",
  "ELEM01" = "HAB",
  "GH" = "HAB",
  "HB" = "HAB",
  "JB" = "HAB",
  "KEYS01" = "HAB",
  "KEYS03" = "HAB",
  "KP01" = "HAB",
  "LC01" = "HAB",
  "LPTNT" = "HAB",
  "LS" = "HAB",
  "LS2" = "HAB",
  "LUC01" = "HAB",
  "RED01" = "HAB",
  "RODS" = "HAB",
  "SBMMEL01" = "HAB",
  "SHADY01" = "HAB",
  "UBL" = "HAB",
  "CL-1" = "HAB",
  "CL-3" = "HAB",
  "CL-4" = "HAB",
  "CL-5" = "HAB",
  "LA-03" = "HAB",
  "NR-02" = "HAB",
  "OA-04" = "HAB",
  "UA-01" = "HAB",
  "UA-06" = "HAB",
  "UA-07" = "HAB",
  "UA-08" = "HAB",
  "PILLS01" = "HAB", #not in cdx
  "LAKEPILS01" = "HAB" #not in cdx
  )

unit_lookup <- c(
  "Temperature, water" = "deg C", 
  "Specific conductance" = "mS/cm", 
  "Resistivity" = "KOhm-cm", 
  "Salinity" = "ppt", 
  "Total dissolved solids" = "g/L", 
  "Dissolved oxygen saturation" = "%",
  "Dissolved oxygen (DO)" = "mg/L", 
  "pH" = "None", 
  "Turbidity" = "NTU",
  "Oil & Grease (HEM)" = "mg/L",
  "Nitrate + Nitrite as N" = "mg/L",
  "Phosphorus, total" = "mg/L",
  "Total Organic Carbon" = "mg/L")

method_id_lookup <- c(
  "SM9223B" = "9223-B",
  "EPA 300.0" = "300.0",
  "ELISA" = "520060",
  "QPCR" = "1611",
  "EPA 1664A" = "1664A",
  "SM4500-NO3 E" = "4500-NO3(E)",
  "SM4500-P F" = "4500-P-F",
  "SM5310C" = "5310-C"
)

method_context_lookup <- c(
  "SM9223B" = "APHA",
  "EPA 300.0" = "USEPA",
  "ELISA" = "ABRAXIS LLC",
  "QPCR" = "USEPA",
  "EPA 1664A" = "USEPA",
  "SM4500-NO3 E" = "APHA",
  "SM4500-P F" = "APHA",
  "SM5310C" = "APHA"
  
)

characteristic_lookup <- c(
  "Oil & Grease (HEM)" = "Oil and Grease",
  "Phosphorus, total" = "Phosphorus",
  "Total Organic Carbon" = "Organic carbon",
  "E. Coli" = "Escherichia coli",
  "Nitrate + Nitrite as N" = "Nitrate + Nitrite",
  "Total Kjeldahl Nitrogen" = "Kjeldahol Nitrogen"
)

method_speciation_lookup <- c(
  "Nitrate + Nitrite as N" = "as N",
  "Nitrate as N" = "as N",
  "Nitrite as N" = "as N",
  "Phosphorus, total" = "as P",
  "Total Nitrogen" = "as N",
  "Total Kjeldahl Nitrogen" = "as N",
  "Orthophosphate" = "as P"
)

```

## Load Data

Click on browse and find the appropriate Alpha Lab excel files on your computer. Select the file that needs to be transformed to WQX format. 

We will read the excel files using the function `read_excel()` from the readxl package. The loading file script will also parse the 'SAMPLENAME' column for the sample location in the column.

Example:
Here is an example of reading in the file using the `read_excel()` function.

```{r, eval=FALSE, class.source = 'fold-show'}
alpha_labs_data <- readxl::read_excel("path/to/alpha-labs-raw-data.xlsx")
```

The chunk below will parse the file. Once the parsing is finished, the data frame will appear below this chunk.

```{r, echo=FALSE}
fileInput("alpha_files",
          "Choose AlphaLab Excel File",
          multiple = FALSE)

all_files <- reactive({
  
  req(input$alpha_files)
  
  raw_alpha_lab <- purrr::map_df(input$alpha_files$datapath, read_excel) |> 
    tidyr::separate(SAMPLENAME, into=c("sample1", "sample2"), sep = " ") |> 
    dplyr::mutate(SAMPLENAME = ifelse(sample1 %in% names(project_id_lookup), sample1, sample2)) |>
    dplyr::select(-sample1, -sample2) |> 
    dplyr::relocate(SAMPLENAME, .before = LABSAMPID)
  
  return (raw_alpha_lab)
})

unique_analytes <- reactive({
  req(input$alpha_files)
  print(unique(all_files()$ANALYTE))
  
})

renderDataTable(
  all_files(), 
  options=list(scrollX=T)
)

```



## QA/QC

Example code:

This chunk shows the `ifelse()` function to test the Project column. The returned value will be either yes or no depending on whether the test is TRUE or FALSE. That value will be saved as the result variable.
```{r, eval=FALSE, class.source = 'fold-show'}
result <- ifelse(alpha_labs_data$PROJECT == "Storm Water Monitoring", "YES", "NO")
```


Here are a few basic QA/QC processes to make sure that the data is in the right format:

```{r}
# ADD QA FUNCTION HERE ---------------------------------------------------------------
# an example of qa qc function, it MUST return a result T/F and a description string
check_correct_project_name <- function(d) {
  result <- ifelse(all(all_files()$PROJECT == "Storm Water Monitoring"), TRUE, FALSE)
  desc <- "Check 'Project' is set to 'Storm Water Monitoring' in all records"
  
  return(list(result = result, description=desc))
} 


check_missing_data_in_sample_name <- function() {
  result <- ifelse(any(is.na(all_files()$SAMPLENAME)), FALSE, TRUE)
  
  return(list(result = result, description="Checking for missing data in 'SAMPLENAME' column"))
}


check_results_has_no_missing <- function() {
  result <- ifelse(any(all_files()$Result == "ND"), FALSE, TRUE)
  
  return(list(result = result, description="Checking for no data (ND) in 'Results' column"))
}


check_locations_in_lookup <- function(){
  result <- ifelse(all(all_files()$SAMPLENAME %in% names(project_id_lookup)), TRUE, FALSE)
  
  return(list(result = result, description="Checking for 'Location' column in listed Project Sites"))
}


check_p_no_data <- function(){
   alpha_without_results <- all_files() |>
    filter(all_files()$`Result` == "ND")

   result <- ifelse("Phosphorus, total" %in% alpha_without_results$ANALYTE, FALSE, TRUE)

   return(list(result = result, description = "Checking for no data (ND) in 'Phosphorus, total'"))

}

check_n_no_data <- function(){
   alpha_without_results <- all_files() |>
    filter(all_files()$`Result` == "ND")

   result <- ifelse("Nitrate + Nitrite as N" %in% alpha_without_results$ANALYTE, FALSE, TRUE)

   return(list(result = result, description = "Checking for no data (ND) in 'Nitrate + Nitrite as N'"))

}

check_oil_no_data <- function(){
   alpha_without_results <- all_files() |>
    filter(all_files()$`Result` == "ND")

   result <- ifelse("Oil & Grease (HEM)" %in% alpha_without_results$ANALYTE, FALSE, TRUE)

   return(list(result = result, description = "Checking for no data (ND) in 'Oil & Grease (HEM)'"))

}


check_c_no_data <- function(){
   alpha_without_results <- all_files() |>
    filter(all_files()$`Result` == "ND")

   result <- ifelse("Total Organic Carbon" %in% alpha_without_results$ANALYTE, FALSE, TRUE)

   return(list(result = result, description = "Checking for no data (ND) in 'Total Organic Carbon'"))

}


check_n_range <- function() {
  
  alpha_with_results <- all_files() |> 
    filter(all_files()$`Result` != "ND")
  
  
  if ("Nitrate + Nitrite as N" %in% alpha_with_results$ANALYTE){
    alpha_with_results <- alpha_with_results |> filter(alpha_with_results$`ANALYTE` == "Nitrate + Nitrite as N")
    result <- with(
      alpha_with_results,
      all(as.numeric(alpha_with_results$Result) < 10 & as.numeric(alpha_with_results$Result) > 0))
    
    return(
      list(result = result, description = "Checking for 'Nitrate and Nitrite as N' smaller than 10 mg/L and bigger than 0 mg/L")
    )
  }
  
}


check_oil_range <- function() {
  
  alpha_with_results <- all_files() |> 
    filter(all_files()$`Result` != "ND")
  
  if ("Oil & Grease (HEM)" %in% alpha_with_results$ANALYTE){
    alpha_with_results <- alpha_with_results |> filter(alpha_with_results$`ANALYTE` == "Oil & Grease (HEM)")
                                                       
    result <- with(
      alpha_with_results,
      all(as.numeric(alpha_with_results$Result) < 3000 & as.numeric(alpha_with_results$Result) > 0))
    
    return(
      list(result = result, description = "Checking for 'Oil & Grease' smaller than 3000 mg/L and bigger than 0 mg/L")
    )
  }
}

check_p_range <- function() {
  
  alpha_with_results <- all_files() |> 
    filter(all_files()$`Result` != "ND")
  
  
  if ("Phosphorus, total" %in% alpha_with_results$ANALYTE){
    alpha_with_results <- alpha_with_results |> filter(alpha_with_results$`ANALYTE` == "Phosphorus, total")
    result <- with(
      alpha_with_results,
      all(as.numeric(alpha_with_results$Result) < 2 & as.numeric(alpha_with_results$Result) > 0))
    
    return(
      list(result = result, description = "Checking for 'Phosphorus, total' smaller than 2 mg/L and bigger than 0 mg/L")
    )
  }
  
}

check_c_range <- function() {
  
  alpha_with_results <- all_files() |> 
    filter(all_files()$`Result` != "ND")
  
  
  if ("Total Organic Carbon" %in% alpha_with_results$ANALYTE){
    alpha_with_results <- alpha_with_results |> filter(alpha_with_results$`ANALYTE` == "Total Organic Carbon")
    result <- with(
      alpha_with_results,
      all(as.numeric(alpha_with_results$Result) < 10 & as.numeric(alpha_with_results$Result) > 0))
    
    return(
      list(result = result, description = "Checking for 'Total Organic Carbon' smaller than 10 mg/L and bigger than 0 mg/L")
    )
  }
  
}

check_tot_coli_range <- function() {
  
  alpha_with_results <- all_files() |> 
    filter(all_files()$`Result` != "ND")
  
  
  if ("Total Coliform" %in% alpha_with_results$ANALYTE){
    alpha_with_results <- alpha_with_results |> filter(alpha_with_results$`ANALYTE` == "Total Coliform")
    result <- with(
      alpha_with_results,
      all(as.numeric(alpha_with_results$Result) < 300000 & as.numeric(alpha_with_results$Result) > 0))
    
    return(
      list(result = result, description = "Checking for 'Total Coliform' smaller than 300000 MPN/100mL and bigger than 0 MPN/100mL")
    )
  }
  
}

check_fec_coli_range <- function() {
  
  alpha_with_results <- all_files() |> 
    filter(all_files()$`Result` != "ND")
  
  
  if ("Fecal Coliform" %in% alpha_with_results$ANALYTE){
    alpha_with_results <- alpha_with_results |> filter(alpha_with_results$`ANALYTE` == "Fecal Coliform")
    result <- with(
      alpha_with_results,
      all(as.numeric(alpha_with_results$Result) < 300000 & as.numeric(alpha_with_results$Result) > 0))
    
    return(
      list(result = result, description = "Checking for 'Fecal Coliform' smaller than 300000 MPN/100mL and bigger than 0 MPN/100mL")
    )
  }
  
}

check_kn_range <- function() {
  
  alpha_with_results <- all_files() |> 
    filter(all_files()$`Result` != "ND")
  
  
  if ("Total Kjeldahl Nitrogen" %in% alpha_with_results$ANALYTE){
    alpha_with_results <- alpha_with_results |> filter(alpha_with_results$`ANALYTE` == "Total Kjeldahl Nitrogen")
    result <- with(
      alpha_with_results,
      all(as.numeric(alpha_with_results$Result) < 10 & as.numeric(alpha_with_results$Result) > 0))
    
    return(
      list(result = result, description = "Checking for 'Total Kjeldahl Nitrogen' smaller than 10 mg/L and bigger than 0 mg/L")
    )
  }
  
}
```


```{r, echo=FALSE}

# function calls all qaqc functions and creates html table data for them
apply_qaqc <- function(qaqc_list) {
  
  purrr::map(qaqc_list, function(f) {
    qaqc_result <- f()
    emoji <- ifelse(qaqc_result$result, emo::ji("check"), emo::ji("x"))
    tagList(
      tags$tr(
      tags$td(paste(qaqc_result$description), style="padding:5px;"),
      tags$td(emoji, style="padding:5px;")
      )
    )
  })
  
}


  # populate this list with all qaqc funcs to apply

qaqc_funcs <- list(
  check_correct_project_name, 
  check_locations_in_lookup,
  check_results_has_no_missing,
  # check_n_no_data,
  check_n_range,
  # check_p_no_data,
  check_p_range,
  # check_c_no_data,
  check_c_range,
  # check_oil_no_data,
  check_oil_range,
  check_fec_coli_range,
  check_tot_coli_range,
  check_kn_range
)

renderUI(
  tags$html(
    tags$table(
      class = "table",
      border = "1",
      style="border:2px solid black;width:70%",
      tags$tr(apply_qaqc(qaqc_funcs)) # <--------------- qaqc get applied here
    )
  )
)


```

## Make Activity Function

Before transforming the results data into WQX format, we will create a function that will help to make the 'Activity ID' column of the WQX file. To make the Activity ID, it needs location ID, date, activity type, equipment name, depth, time and equipment comment. Run this function with a `mutate()` statement to create the 'Activity ID.'


```{r}
make_activity_id <-
  function(location_id,
           date,
           activity_type,
           equipment_name,
           depth = NULL,
           time = NULL,
           equipment_comment = NULL) {
    YYYYMMDD <- gsub('/', '', date)
    activity <- ifelse(activity_type == "Sample-Routine", "SR", "FM")
    equipment <- case_when(
      equipment_name == "Probe/Sensor" ~ "PS",
      equipment_name == "Water Bottle" ~ "WB",
      TRUE ~ NA_character_
    )
    hhmm <- gsub(':', '', time)
    equipment_comment <- case_when(
      equipment_comment == "Hydrolab Surveyor DS5 Multiprobe" ~ "Hydro",
      equipment_comment == "AlgaeChek Ultra Fluorometer" ~ "Algae",
      TRUE ~ NA_character_
    )
    depth <- ifelse(is.na(depth), "", depth)
    paste(location_id,
          YYYYMMDD,
          hhmm,
          activity,
          equipment,
          depth,
          equipment_comment,
          sep = ":")
  }
```

## Data Transformation

To transform the data into WQX format, we will create columns based on the WQX schema using the `mutate()` function from the tidyverse package. We will then drop the columns we do not need and reorder the data.

One crucial function is the `casewhen()` function. This function lets the user create multiple conditions; if this condition is met, the column will be a specified value. If it is not met, it is a different value.

For example for the following code: `all_files <- all_files() |> mutate("Characteristic Name" = case_when( ANALYTE == "E. Coli" ~ "Escherichia coli",` `ANALYTE == "Oil & Grease (HEM)" ~ "Oil and Grease", ANALYTE == "Phosphorus, total" ~ "Phosphorus",` `ANALYTE == "Total Organic Carbon" ~ "Organic carbon", TRUE ~ ANALYTE)`

It is paraphrased as we create a new column named 'Characteristic Name' in the data frame all_files. If the cell value from the 'ANALYTE' column from all_files is E. Coli (the condition), the corresponding cell value in the 'Characteristic Name' column will be "Escherichia coli". If the cell value from the 'ANALYTE' column is "Oil & Grease (HEM)," the 'Characteristic name' column will be "Oil and Grease ."If the cell value from the 'ANALYTE' column is "Total Organic Carbon," the 'Characteristic name' column will be "Organic carbon ." And if the 'ANALYTE' column does not match any of these conditions, fill the rest of the 'Characteristic Name' column with values from the 'ANALYTE' column.

The `ifelse()` function works similarly, except it only takes in one condition. If that condition is met, the new column will be a new value. If it is not met, it will be another value.

Example code:
This chunk creates a new column called "Project ID", with its value being "SW".
```{r, eval=FALSE, class.source = 'fold-show'}
alpha_labs_data <- alpha_labs_data |> 
  mutate("Project ID" = "SW")
```

```{r}
partial_clean_alpha_lab <- reactive({
  req(input$alpha_files)
  
  all_files <- all_files() |>
    mutate(
      "Project ID" = project_id_lookup[SAMPLENAME],
      "Monitoring Location ID" = SAMPLENAME,
      "Activity ID User Supplied (PARENTs)" = NA,
      "Activity Type" = "Sample-Routine",
      "Activity Media Name" = MATRIX,
      "Activity Start Date" = format(mdy_hms(SAMPDATE), "%m/%d/%Y"),
      "Activity Start Time" = format(mdy_hms(SAMPDATE), "%H:%M"),
      "Activity Start Time Zone" = "PST",
      "Activity Depth/Height Measure" = "0.152",
      "Activity Depth/Height Unit" = "m",
      "Sample Collection Method ID" = "BVR SWQAPP",
      "Sample Collection Method Context" = "CA_BVR",
      "Sample Collection Equipment Name" = "Water Bottle",
      "Sample Collection Equipment Comment" = NA,
      "Characteristic Name" = ifelse(
        ANALYTE %in% names(characteristic_lookup), characteristic_lookup[ANALYTE], ANALYTE),
      "Characteristic Name User Supplied" = NA,
      "Method Speciation" = ifelse(
        ANALYTE %in% names(method_speciation_lookup), method_speciation_lookup[ANALYTE], NA),
      "Result Detection Condition" = case_when(
        Result == "ND" ~ "Not Detected",
        Result == "Absent" ~ "Not Present",
        Result == "Present" ~ "Detected Not Quantified",
        TRUE ~ NA_character_
      ),
      "Result Value" = ifelse(
        Result == "Absent" |
          Result == "ND" | Result == "Present",
        NA_character_,
        Result
      ),
      "Result Unit" = ifelse(UNITS == ".", NA_character_, UNITS),
      "Result Measure Qualifier" = NA,
      "Result Sample Fraction" = "Total",
      "Result Status ID" = "Final",
      "ResultTemperatureBasis" = NA,
      "Statistical Base Code" = NA,
      "ResultTimeBasis" = NA,
      "Result Value Type" = ifelse(is.na(Result), NA_character_, "Actual"),
      "Result Analytical Method ID" = ifelse(
        is.na(METHODNAME), NA_character_, method_id_lookup[METHODNAME]),
      "Activity ID (CHILD-subset)" = make_activity_id(
        location_id = `Monitoring Location ID`,
        date = `Activity Start Date`,
        time = `Activity Start Time`,
        activity_type = `Activity Type`,
        equipment_name = `Sample Collection Equipment Name`,
        depth = `Activity Depth/Height Measure`
      ),
      "Result Analytical Method Context" = method_context_lookup[METHODNAME],
      "Analysis Start Date" = format(mdy_hms(ANADATE), "%m/%d/%Y"),
      "Result Detection/Quantitation Limit Type" = ifelse(DL == "NA", NA_character_, "Method Detection Level"),
      "Result Detection/Quantitation Limit Measure" = ifelse(DL == "NA", NA_character_, DL),
      "Result Detection/Quantitation Limit Unit" = ifelse(UNITS == ".", NA_character_, UNITS),
      "Result Comment" = NA
      
    ) %>%
    select(-c(0:48)) %>%
    relocate("Activity ID (CHILD-subset)", .before = "Activity ID User Supplied (PARENTs)")
})

```

One last step that we need to do before getting a WQX ready dataset is to make sure that both 'Result Unit' and 'Result Analytical Method Context' column is empty if the 'Result Analytical Method ID' and 'Result Unit' columns are respectively empty. We also have to make sure that the columns are in the right order (using the function `relocate()` to move the columns).

Example code:
In this chunk we move the column "Result Analytical Method Context" to before the column "Analysis Start Date".
```{r, eval=FALSE, class.source = 'fold-show'}
alpha_labs_data <- alpha_labs_data |> 
  relocate("Result Analytical Method Context", .before = "Analysis Start Date")
```

The rendered table shows the WQX ready dataframe.

```{r}
clean_alpha_lab <- reactive({
  req(input$alpha_files)
  
  partial_clean_alpha_lab <-  partial_clean_alpha_lab() |>
    mutate(
      "Result Analytical Method Context" = ifelse(
        is.na(partial_clean_alpha_lab()$`Result Analytical Method ID`),
        NA_character_,
        partial_clean_alpha_lab()$`Result Analytical Method Context`
      ),
      "Result Unit" = ifelse(
        is.na(partial_clean_alpha_lab()$`Result Value`),
        NA_character_,
        partial_clean_alpha_lab()$`Result Unit`
      )
    ) |>
    relocate("Result Analytical Method Context", .before = "Analysis Start Date")
})



renderDataTable(
  clean_alpha_lab(),
  options=list(scrollX=T)
)

```

## Data Download

Lastly, we export the data to a CSV file in the data folder. Cells with NAs are saved as blank cells. 

Click on the Download button for the WQX-ready CSV file. 

```{r, echo=FALSE}
unique_signature <- format(lubridate::now(), "%Y%m%d_%H%M%S")

downloadHandler(
  filename = function() {
    paste("alpha-lab-data-", unique_signature, ".csv", sep = "")
  },
  content = function(file) {
    write.csv(clean_alpha_lab(),
              file,
              na = '',
              row.names = FALSE)
  }
)


```

## Data Upload To CDX-WQX

The 'Upload WQX File Path' is automatically populated with the file path of the downloaded CSV file from the previous step. Please double-check and make sure the file is in the default Downloads folder. If it is not, enter the full location of the file.

The 'CDX Account Info Path' is also automatically populated with the file path of the CSV file containing your WQX API Key, Username, and Config ID. Please double check the file is in the big_valley_wax folder.  

Use the 'Upload To WQX' button below to directly upload the previously downloaded data to CDX. It will take a few seconds for the upload status to show up. Once it does, you may close the markdown. CDX will send an email to the user with the final status once the upload finishes.

```{r}
path_to_download <- file.path(Sys.getenv("USERPROFILE"), "Downloads")
file_path <-
  str_replace_all(
    paste(
      path_to_download,
      "/alpha-lab-data-",
      unique_signature,
      ".csv",
      sep = ""
    ),
    "\\\\",
    "/"
  )
file_path_to_cdx_account <- paste0(getwd(), "/cdx-account-info.csv")
textInput("wqx_upload", "Upload WQX File Path", value = file_path)
textInput("cdx_account_info_file", "CDX Account Info File Path", value = file_path_to_cdx_account)
actionButton("button", "Upload To WQX")

conditionalPanel(condition="$('html').hasClass('shiny-busy')",
                            tags$div(HTML("<b> Starting WQX upload. Please wait 25 seconds for the upload status from CDX...</b>"),id="loadmessage"))

upload_data_resp <- eventReactive(input$button, {
  login_info <- read_csv(input$cdx_account_info_file)
  
  API_KEY = login_info$WQX_API_KEY
  USER_ID = login_info$USER_ID
  CONFIG_ID = login_info$CONFIG_ID
  FILE_PATH = input$wqx_upload
  FILE_NAME =  paste("alpha-lab-data-", unique_signature, ".csv", sep = "")
  
  session <- cdx(USER_ID, API_KEY, FILE_PATH, FILE_NAME)
  file_id <- cdx_upload(session = session)
  dataset_id <-
    cdx_import(
      session = session,
      file_id = file_id,
      config_id = CONFIG_ID,
      params = c("newOrExistingData", "0")
    )

  Sys.sleep(25)
  status <- cdx_get_status(session, dataset_id)
  
  return(status$StatusName)
})

renderUI(
  if (upload_data_resp() == "Import Failed") {
  tags$p(tags$b("Import failed."), "Please retry upload.")
} else
  {
    tags$p(
      tags$b("Application is importing data onto CDX."),
      tags$br(),
      "You may now close this document. Check email or CDX website for the final upload confirmation."
    )
  }
)

```





