---
title: "Alpha Lab Data Cleaning and Upload"
author: "Inigo Peng"
date: "2022-10-12"
runtime: shiny
output: 
  html_document:
    code_folding: hide
---
## Description of Data

**Collection method:**

Sampling Bottles

**Raw File format:**

AlphaLab CSV lab report

**Monitoring (Project ID)**

Storm Water Sampling (SW); Tap Water Sampling (CalWATCH).

**WQX submittal required?**

Storm Water Sampling: YES; Tap Water Sampling: NO.

**Note:**

This script is a dynamic markdown where the user will be able to interact with the document. To run the document, click on "Run Document" to render it into an HTML format. 

Usually, to run a markdown script, the user needs to make sure the project is set to the appropriate working directory. However, since this is an interactive markdown, we do not have to worry about that. The only files we need are the Alpha Lab excel files that needs to get updated.

In these markdown documents, to see the code that are used to process the data, click on the "code" button on the right. If you have any question about any of the functions used, type in ? followed by the function (eg. `?mutate` in the console,  or google search the function to see detailed documentation on the function.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      tidy.opts = list(width.cutoff = 60), tidy = TRUE)

library(tidyverse)
library(readxl)
library(lubridate)
library(wqxWeb)
library(DT)
library(shiny)
library(stringr)
```
```{r, include=FALSE}
project_id_lookup <- c(
  "M1" = "MS",
  "M2" = "MS",
  "M3" = "MS",
  "M4" = "MS",
  "HSP" = "SW",
  "BVSWD1" = "SW",
  "NBPRSC" = "SW",
  "RSTCC" = "SW",
  "BVCL1" = "SW",
  "BVRTC1" = "SW",
  "BVRTCC" = "SW",
  "BVSWDRV" = "SW",#not in cdx
  "RVSI1" = "SW",#not in cdx
  "RVSI2" = "SW",#not in cdx
  "BVCL2" = "CLM",	
  "BVCL3" = "CLM",	
  "BVCL5" = "CLM",	
  "BVCL6" = "CLM",	
  "BVCL11" = "CLM",	
  "BVCL12" = "CLM",	
  "BVCL13" = "CLM",	
  "BVCL14" = "CLM",	
  "BVCL15" = "CLM",	
  "BVCL16" = "CLM",	
  "BVCL17" = "CLM",	
  "BVCL18" = "CLM",	
  "BVCL19" = "CLM",	
  "BVCL20" = "CLM",
  "FC1" = "CS", 
  "FC2" = "CS",
  "FC3" = "CS",#not in cdx
  "MC1" = "CS",
  "MC2" = "CS",
  "TC1" = "CS",
  "AC1" = "CS",
  "AC2" = "CS",
  "AC3" = "CS",
  "AC4" = "CS",
  "MCC1" = "CS",
  "KC1" = "CS",
  "CC1" = "CS",
  "SC1" = "CS",
  "SC2" = "CS",#not in cdx
  "SHC2"= "CS",#not in cdx
  "CC2" = "CS",#not in cdx
  "SIEG01" = "CS",#not in cdx
  "COOP01" = "CS",#not in cdx
  "CLOV01" = "CS",#not in cdx
  "DRY01" = "CS",#not in cdx
  "COY01" = "CS",#not in cdx
  "SCOT01"= "CS",#not in cdx
  "AND01" = "CS",#not in cdx
  "NFORK01" = "CS",#not in cdx
  "LONG01" = "CS", #not in cdx
  "PUT01" = "CS", #not in cdx
  "MID01"= "CS", #not in cdx
  "KC5"= "CS", #not in cdx
  "AP01" = "HAB",
  "BP" = "HAB",
  "CLOAKS01" = "HAB",
  "CLV7" = "HAB",
  "CP" = "HAB",
  "ELEM01" = "HAB",
  "GH" = "HAB",
  "HB" = "HAB",
  "JB" = "HAB",
  "KEYS01" = "HAB",
  "KEYS03" = "HAB",
  "KP01" = "HAB",
  "LC01" = "HAB",
  "LPTNT" = "HAB",
  "LS" = "HAB",
  "LS2" = "HAB",
  "LUC01" = "HAB",
  "RED01" = "HAB",
  "RODS" = "HAB",
  "SBMMEL01" = "HAB",
  "SHADY01" = "HAB",
  "UBL" = "HAB",
  "CL-1" = "HAB",
  "CL-3" = "HAB",
  "CL-4" = "HAB",
  "CL-5" = "HAB",
  "LA-03" = "HAB",
  "NR-02" = "HAB",
  "OA-04" = "HAB",
  "UA-01" = "HAB",
  "UA-06" = "HAB",
  "UA-07" = "HAB",
  "UA-08" = "HAB",
  "PILLS01" = "HAB", #not in cdx
  "LAKEPILS01" = "HAB" #not in cdx
  )

unit_lookup <- c(
  "Temperature, water" = "deg C", 
  "Specific conductance" = "mS/cm", 
  "Resistivity" = "KOhm/cm", 
  "Salinity" = "ppt", 
  "Total dissolved solids" = "g/L", 
  "Dissolved oxygen saturation" = "%",
  "Dissolved oxygen (DO)" = "mg/L", 
  "pH" = "None", 
  "Turbidity" = "NTU",
  "Oil & Grease (HEM)" = "mg/L",
  "Nitrate + Nitrite as N" = "mg/L",
  "Phosphorus, total" = "mg/L",
  "Total Organic Carbon" = "mg/L")

method_id_lookup <- c(
  "SM9223B" = "9223-B",
  "EPA 300.0" = "300.0",
  "ELISA" = "520060",
  "QPCR" = "1611",
  "EPA 1664A" = "1664A",
  "SM4500-NO3 E" = "4500-NO3(E)",
  "SM4500-P F" = "4500-P-F",
  "SM5310C" = "5310-C"
)

method_context_lookup <- c(
  "SM9223B" = "APHA",
  "EPA 300.0" = "USEPA",
  "ELISA" = "ABRAXIS LLC",
  "QPCR" = "USEPA",
  "EPA 1664A" = "USEPA",
  "SM4500-NO3 E" = "APHA",
  "SM4500-P F" = "APHA",
  "SM5310C" = "APHA"
  
)

characteristic_lookup <- c(
  "Oil & Grease (HEM)" = "Oil and Grease",
  "Phosphorus, total" = "Phosphorus",
  "Total Organic Carbon" = "Organic carbon",
  "E. Coli" = "Escherichia coli"
)

project_sites <- c(
  "M1",
  "M2",
  "M3",
  "M4",
  "HSP",
  "BVSWD1",
  "NBPRSC",
  "RSTCC",
  "BVCL1",
  "BVRTC1",
  "BVRTCC",
  "BVSWDRV",#not in cdx
  "RVSI1",#not in cdx
  "RVSI2",#not in cdx
  "BVCL2",	
  "BVCL3",	
  "BVCL5",	
  "BVCL6",	
  "BVCL11",	
  "BVCL12",	
  "BVCL13",	
  "BVCL14",	
  "BVCL15",	
  "BVCL16",	
  "BVCL17",	
  "BVCL18",	
  "BVCL19",	
  "BVCL20",
  "FC1",
  "FC2",
  "FC3",#not in cdx
  "MC1",
  "MC2",
  "TC1",
  "AC1",
  "AC2",
  "AC3",
  "AC4",
  "MCC1",
  "KC1",
  "CC1",
  "SC1",
  "SC2",#not in cdx
  "SHC2",#not in cdx
  "CC2",#not in cdx
  "SIEG01",#not in cdx
  "COOP01",#not in cdx
  "CLOV01",#not in cdx
  "DRY01",#not in cdx
  "COY01",#not in cdx
  "SCOT01",#not in cdx
  "AND01",#not in cdx
  "NFORK01",#not in cdx
  "LONG01", #not in cdx
  "PUT01", #not in cdx
  "MID01", #not in cdx
  "KC5", #not in cdx
  "AP01",
  "BP",
  "CLOAKS01",
  "CLV7",
  "CP",
  "ELEM01",
  "GH",
  "HB",
  "JB",
  "KEYS01",
  "KEYS03",
  "KP01",
  "LC01",
  "LPTNT",
  "LS",
  "LS2",
  "LUC01",
  "RED01",
  "RODS",
  "SBMMEL01",
  "SHADY01",
  "UBL",
  "CL-1",
  "CL-3",
  "CL-4",
  "CL-5",
  "LA-03",
  "NR-02",
  "OA-04",
  "UA-01",
  "UA-06",
  "UA-07",
  "UA-08",
  "PILLS01", #not in cdx
  "LAKEPILS01"  #not in cdx
)
```


## Load Data

Click on browse and find the appropriate Alpha Lab excel files on your computer. Select all of the files (could select multiple) that need to be transformed to wqx format. 

We will read the excel files using the function `read_excel()` from the readxl package. The script will create a dataframe of AlphaLab data by binding each file by row using the function `rbind()`. The end result is one data frame with all of the data from the folder.

Lastly, the script will parse the 'SAMPLENAME' column for the sample location in the column.

Once the parsing is finished, the first five rows of the AlphaLab dataframe will appear below this chunk.

```{r, echo=FALSE}
fileInput("alpha_files",
          "Choose AlphaLab Excel File",
          multiple = TRUE)

all_files <- reactive({
  
  req(input$alpha_files)
  
  raw_alpha_lab <-
    map_df(purrr::map(input$alpha_files$datapath, read_excel), rbind)
  for (row in 1:nrow(raw_alpha_lab)) {
    for (n in 1:length(strsplit(raw_alpha_lab$SAMPLENAME, " ")[[row]])) {
      if (strsplit(raw_alpha_lab$SAMPLENAME, " ")[[row]][n] %in% project_sites) {
        raw_alpha_lab$SAMPLENAME[row] = strsplit(raw_alpha_lab$SAMPLENAME, " ")[[row]][n]
      }
    }
  }
  return (raw_alpha_lab)
})

renderDataTable(
  all_files(), 
  options=list(scrollX=T)
)

```

## QA/QC

Here are a few basic QA/QC processes to make sure that the data is in the right format:


```{r}
renderUI({
  HTML(
    paste(
      "<ul><li> Are all values in the 'Project' column 'Storm Water Monitoring': ",
      "<b>",
      all(all_files()$PROJECT == "Storm Water Monitoring"),
      "</b></li></ul>"
    )
  )
})

#Check if SAMPLENAME is N
renderUI({
  HTML(
    paste(
      "<ul><li> Are there any NA values in the 'SAMPLENAME' column: ",
      "<b>",
      any(is.na(all_files()$SAMPLENAME)),
      "</b></li></ul>"
    )
  )
})

#Check for all unique analytes
renderUI({
  HTML(
    paste(
      "<ul><li> Unique water quality analytes collected in the 'ANALYTE' column: ",
      "<b>",
      paste(unique(all_files()$ANALYTE, "\n")),
      "</b>",
      "</ul></li>"
    )
  )
})

```

## Make Activity Function

Before transforming the results data into WQX format, we will create a function that will help to make the 'Activity ID' column of WQX file. To make the Activity ID, it needs location ID, date, activity type, equipment name, depth, time and equipment comment. Simply run this function with a `mutate()` statement to create the 'Activity ID'.


```{r}
make_activity_id <-
  function(location_id,
           date,
           activity_type,
           equipment_name,
           depth = NULL,
           time = NULL,
           equipment_comment = NULL) {
    YYYYMMDD <- gsub('/', '', date)
    activity <- ifelse(activity_type == "Sample-Routine", "SR", "FM")
    equipment <- case_when(
      equipment_name == "Probe/Sensor" ~ "PS",
      equipment_name == "Water Bottle" ~ "WB",
      TRUE ~ NA_character_
    )
    hhmm <- gsub(':', '', time)
    equipment_comment <- case_when(
      equipment_comment == "Hydrolab Surveyor DS5 Multiprobe" ~ "Hydro",
      equipment_comment == "AlgaeChek Ultra Fluorometer" ~ "Algae",
      TRUE ~ NA_character_
    )
    depth <- ifelse(is.na(depth), "", depth)
    paste(location_id,
          YYYYMMDD,
          hhmm,
          activity,
          equipment,
          depth,
          equipment_comment,
          sep = ":")
  }
```

## Data Transformation

To transform the data into WQX format, we will create columns based on WQX schema using the `mutate()` function from the tidyverse package. We will then drop the columns we do not need and reorder the data.

One important function is the `casewhen()` function. This function lets the user create multiple conditions, and if this condition is met, the column will be a specified value. If it is not met, it is a different value.

For example, in the code `all_files <- all_files() |> mutate("Characteristic Name" = case_when( ANALYTE == "E. Coli" ~ "Escherichia coli",` `ANALYTE == "Oil & Grease (HEM)" ~ "Oil and Grease", ANALYTE == "Phosphorus, total" ~ "Phosphorus",` `ANALYTE == "Total Organic Carbon" ~ "Organic carbon", TRUE ~ ANALYTE)`

It could be read as: we are creating a new column named 'Characteristic Name' in the dataframe all_files. If the cell value from the 'ANALYTE' column from all_files is E. Coli (the condition), the corresponding cell value in the 'Characteristic Name' column will be "Eschericia coli". If the cell value from 'ANALYTE' column is "Oil & Grease (HEM)", the 'Characteristic name' column will be "Oil and Grease". If the cell value from 'ANALYTE' column is "Total Organic Carbon", the 'Characteristic name' column will be "Organic carbon".  And if the 'ANALYTE' column does not match any of these conditions, fill the rest of the 'Characteristic Name' column with values from the 'ANALYTE' column.

The `ifelse()` function works in the same way except it only takes in one condition. If that condition is met, the new column will be a new value. If it is not met, it will be another value.


```{r}
partial_clean_alpha_lab <- reactive({
  req(input$alpha_files)
  
  all_files <- all_files() |>
    mutate(
      "Project ID" = "SW",
      "Monitoring Location ID" = SAMPLENAME,
      "Activity ID User Supplied (PARENTs)" = NA,
      "Activity Type" = "Sample-Routine",
      "Activity Media Name" = MATRIX,
      "Activity Start Date" = format(mdy_hms(SAMPDATE), "%m/%d/%Y"),
      "Activity Start Time" = format(mdy_hms(SAMPDATE), "%H:%M"),
      "Activity Start Time Zone" = "PST",
      "Activity Depth/Height Measure" = "0.152",
      "Activity Depth/Height Unit" = "m",
      "Sample Collection Method ID" = "BVR SWQAPP",
      "Sample Collection Method Context" = "CA_BVR",
      "Sample Collection Equipment Name" = "Water Bottle",
      "Sample Collection Equipment Comment" = NA,
      "Characteristic Name" = case_when(
        ANALYTE == "E. Coli" ~ "Escherichia coli",
        ANALYTE == "Oil & Grease (HEM)" ~ "Oil and Grease",
        ANALYTE == "Phosphorus, total" ~ "Phosphorus",
        ANALYTE == "Total Organic Carbon" ~ "Organic carbon",
        TRUE ~ ANALYTE
      ),
      "Characteristic Name User Supplied" = NA,
      "Method Speciation" = case_when(
        ANALYTE == "Nitrate as N" ~ "as N",
        ANALYTE == "Phosphorus, total" ~ "as P",
        ANALYTE == "Nitrate + Nitrite as N" ~ "as N",
        TRUE ~ NA_character_
      ),
      "Result Detection Condition" = case_when(
        Result == "ND" ~ "Not Reported",
        Result == "Absent" ~ "Not Present",
        Result == "Present" ~ "Detected Not Quantified",
        TRUE ~ NA_character_
      ),
      "Result Value" = ifelse(
        Result == "Absent" |
          Result == "ND" | Result == "Present",
        NA_character_,
        Result
      ),
      "Result Unit" = ifelse(UNITS == ".", NA_character_, UNITS),
      "Result Measure Qualifier" = NA,
      "Result Sample Fraction" = "Total",
      "Result Status ID" = "Final",
      "ResultTemperatureBasis" = NA,
      "Statistical Base Code" = NA,
      "ResultTimeBasis" = NA,
      "Result Value Type" = ifelse(is.na(Result), NA_character_, "Actual"),
      "Result Analytical Method ID" = ifelse(is.na(METHODNAME), NA_character_, method_id_lookup[METHODNAME]),
      "Activity ID (CHILD-subset)" = make_activity_id(
        location_id = `Monitoring Location ID`,
        date = `Activity Start Date`,
        time = `Activity Start Time`,
        activity_type = `Activity Type`,
        equipment_name = `Sample Collection Equipment Name`,
        depth = `Activity Depth/Height Measure`
      ),
      "Result Analytical Method Context" = method_context_lookup[METHODNAME],
      "Analysis Start Date" = format(mdy_hms(ANADATE), "%m/%d/%Y"),
      "Result Detection/Quantitation Limit Type" = ifelse(DL == "NA", NA_character_, "Method Detection Level"),
      "Result Detection/Quantitation Limit Measure" = ifelse(DL == "NA", NA_character_, DL),
      "Result Detection/Quantitation Limit Unit" = ifelse(UNITS == ".", NA_character_, UNITS),
      "Result Comment" = NA
      
    ) %>%
    select(-c(0:48)) %>%
    relocate("Activity ID (CHILD-subset)", .before = "Activity ID User Supplied (PARENTs)")
})

```

One last step that we need to do before getting a WQX ready dataset is to make sure that both 'Result Unit' and 'Result Analytical Method Context' column is empty if the 'Result Analytical Method ID' and 'Result Unit' columns are respectively empty. We also have to make sure that the columns are in the right order (using the function `relocate()` to move the columns).

The rendered table shows the WQX ready dataframe.

```{r}
clean_alpha_lab <- reactive({
  req(input$alpha_files)
  
  partial_clean_alpha_lab <-  partial_clean_alpha_lab() |>
    mutate(
      "Result Analytical Method Context" = ifelse(
        is.na(partial_clean_alpha_lab()$`Result Analytical Method ID`),
        NA_character_,
        partial_clean_alpha_lab()$`Result Analytical Method Context`
      ),
      "Result Unit" = ifelse(
        is.na(partial_clean_alpha_lab()$`Result Value`),
        NA_character_,
        partial_clean_alpha_lab()$`Result Unit`
      )
    ) |>
    relocate("Result Analytical Method Context", .before = "Analysis Start Date")
})



renderDataTable(
  clean_alpha_lab(),
  options=list(scrollX=T)
)

```

## Data Download

Lastly we export the data to a CSV file in the data folder. Cells with NAs are saved as blank cells. 

Click on the Download button for the WQX ready CSV file. 

```{r, echo=FALSE}
unique_signature <- format(lubridate::now(), "%Y%m%d_%H%M%S")

downloadHandler(
  filename = function() {
    paste("alpha-lab-data-", unique_signature, ".csv", sep = "")
  },
  content = function(file) {
    write.csv(clean_alpha_lab(),
              file,
              na = '',
              row.names = FALSE)
  }
)


```

## Data Upload To CDX-WQX

The 'Upload WQX File Path' is automatically populated with the location of the downloaded CSV file from the previous step. Please double check and make sure the file is in the default Downloads folder. If it is not, enter the full location of the file.

Use the 'Upload To WQX' button below to directly upload the previously downloaded data to CDX. It will take a a few seconds for the upload status to show up. Once it does, you may close the markdown. CDX will send an email to the user with the final status once the upload finishes.

```{r}
path_to_download <- file.path(Sys.getenv("USERPROFILE"), "Downloads")
file_path <-
  str_replace_all(
    paste(
      path_to_download,
      "/alpha-lab-data-",
      unique_signature,
      ".csv",
      sep = ""
    ),
    "\\\\",
    "/"
  )

textInput("wqx_upload", "Upload WQX File Path", value = file_path)
actionButton("button", "Upload To WQX")

print_statement <- eventReactive(input$button, {
  print("Starting WQX upload - please wait for upload status")
})

renderPrint({
  print_statement()
})

status <- eventReactive(input$button, {
  API_KEY = Sys.getenv("WQX_API_KEY")
  USER_ID = "INIGOPENG"
  FILE_PATH = input$wqx_upload
  FILE_NAME =  paste("alpha-lab-data-", unique_signature, ".csv", sep = "")
  config_id = 8030
  session <- cdx(USER_ID, API_KEY, FILE_PATH, FILE_NAME)
  file_id <- cdx_upload(session = session)
  dataset_id <-
    cdx_import(
      session = session,
      file_id = file_id,
      config_id = config_id,
      params = c("newOrExistingData", "0")
    )
  # 
  status <- cdx_get_status(session, dataset_id)
  
  return(status$StatusName)

})


renderPrint({
  paste("Upload Status:", status())
  })

```





