---
title: "AlphaLab Data Wrangling"
author: "Inigo Peng"
date: "2022-10-12"
runtime: shiny
output: html_document
---
# AlphaLab Data
## Description of Data

**Collection method:**

Sampling Bottles

**Raw File format:**

AlphaLab CSV lab report

**Monitoring (Project ID)**

Storm Water Sampling (SW), Tap Water Sampling (CalWATCH).

**WQX submittal required?**

Storm Water Sampling: YES. Tap Water Sampling: NO.

**Note:**

This script is a dynamic markdown where the user will be able to interact with the document. To run the document, click on "Run Document" to render it into an HTML format. 

Usually, to run a markdown script, the user needs to make sure the project is set to the appropriate working directory. However, since this is an interactive markdown, we do not have to worry about that. The files we need are the Alpha Lab excel files that needs to get updated, and the lookup_objects.rdata.

In these markdown documents, the shaded sections are the code that are used to process the data. If you have any question about any of the functions used, type in ? followed by the function (eg. ?mutate in the console,  or google search the function to see detailed documentation on the function.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(lubridate)
```

## Data transformations

First we need to load the lookup table named "lookup_object.rdata". This file contains project ID lookup, unit lookup (eg. Temperature, water - deg C), method lookup, method context lookup, and project sites lookup. 

Click on Browse to load the file. 

```{r}
fileInput("lookup_objects", "Load lookup_objects.rdata")

load_file <- reactive({
  req(input$lookup_objects)
  if (is.null(input$lookup_objects))
    return(NULL)
  inFile <- isolate({input$lookup_objects})
  file <- inFile$datapath
  sessionEnvir <- sys.frame()
  # load(file, envir = e)
  # load the file into session environment and get it from there
  load(file, sessionEnvir)
  # project_id_lookup <- e[[name[1]]]
  # # print(data1)
  # unit_lookup <- e[[name[2]]]
  # method_id_lookup <- e[[name[3]]]
  # method_context_lookup <- e[[name[4]]]
  # characteristic_lookup <- e[[name[5]]]
  # project_sites <- e[[name[6]]]
  # data <- sessionEnvir[[name]]
  # return(project_sites)
})

```





Click on browse and find the appropriate Alpha Lab excel files on your computer. Select all of the files (could select multiple) that need to be transformed to wqx format. 

We will read the excel files using the function read_excel() from the readxl package. The script will create a dataframe of AlphaLab data by binding each file by row using the function rbind(). The end result is one data frame with all of the data from the folder.

Lastly, the script will parse the 'SAMPLENAME' column for the sample location in the column.

Once the parsing is finished, the first five rows of the AlphaLab dataframe will appear below this chunk.

```{r}
fileInput("alpha_files",
          "Choose AlphaLab Excel File",
          accept = ".xlsx",
          multiple = TRUE)

all_files <- reactive({
  print(load_file())
  get(load_file()[1])
  # print(df1)
  print(project_sites)
  
  req(input$alpha_files)
  raw_alpha_lab <-
    map_df(purrr::map(input$alpha_files$datapath, read_excel), rbind)
  for (row in 1:nrow(raw_alpha_lab)) {
    for (n in 1:length(strsplit(raw_alpha_lab$SAMPLENAME, " ")[[row]])) {
      if (strsplit(raw_alpha_lab$SAMPLENAME, " ")[[row]][n] %in% project_sites) {
        raw_alpha_lab$SAMPLENAME[row] = strsplit(raw_alpha_lab$SAMPLENAME, " ")[[row]][n]
      }
    }
  }
  return (raw_alpha_lab)
})

renderTable({
  head(all_files())
})

```


Before transforming the results data into WQX format, we will create a function that will help to make the `Activity ID` column of WQX file. To make the Activity ID, it needs location ID, date, activity type, equipment name, depth, time and equipment comment. Simply run this function with a mutate() statement to create the `Activity ID`.


```{r}
make_activity_id <-
  function(location_id,
           date,
           activity_type,
           equipment_name,
           depth = NULL,
           time = NULL,
           equipment_comment = NULL) {
    YYYYMMDD <- gsub('/', '', date)
    activity <- ifelse(activity_type == "Sample-Routine", "SR", "FM")
    equipment <- case_when(
      equipment_name == "Probe/Sensor" ~ "PS",
      equipment_name == "Water Bottle" ~ "WB",
      TRUE ~ NA_character_
    )
    hhmm <- gsub(':', '', time)
    equipment_comment <- case_when(
      equipment_comment == "Hydrolab Surveyor DS5 Multiprobe" ~ "Hydro",
      equipment_comment == "AlgaeChek Ultra Fluorometer" ~ "Algae",
      TRUE ~ NA_character_
    )
    depth <- ifelse(is.na(depth), "", depth)
    paste(location_id,
          YYYYMMDD,
          hhmm,
          activity,
          equipment,
          depth,
          equipment_comment,
          sep = ":")
  }
```

To transform the data into WQX format, we will create columns based on WQX schema using the mutate() function from the tidyverse package. We will then drop the columns we do not need and reorder the data.

One important function is the casewhen() function. This function lets the user create multiple conditions, and if this condition is met, the column will be a specified value. If it is not met, it is a different value.

For example, in the code 

all_files <- all_files() |> 
mutate("Characteristic Name" = case_when(
        ANALYTE == "E. Coli" ~ "Escherichia coli",
        ANALYTE == "Oil & Grease (HEM)" ~ "Oil and Grease",
        ANALYTE == "Phosphorus, total" ~ "Phosphorus",
        ANALYTE == "Total Organic Carbon" ~ "Organic carbon",
        TRUE ~ ANALYTE
      )

It could be read as: we are creating a new column named `Characteristic Name` in the dataframe all_files. If the cell value from the `ANALYTE` column from all_files is E. Coli (the condition), the corresponding cell value in the `Characteristic Name` column will be "Eschericia coli". If the cell value from `ANALYTE` column is "Oil & Grease (HEM)", the `Characteristic name` column will be "Oil and Grease". If the cell value from `ANALYTE` column is "Total Organic Carbon", the `Characteristic name` column will be "Organic carbon".  And if the `ANALYTE` column does not match any of these conditions, fill the rest of the `Characteristic Name` column with values from the `ANALYTE` column.

The ifelse() function works in the same way except it only takes in one condition. If that condition is met, the new column will be a new value. If it is not met, it will be another value.


```{r}
partial_clean_alpha_lab <- reactive({
  req(input$alpha_files)
  
  all_files <- all_files() |>
    mutate(
      "Project ID" = "SW",
      "Monitoring Location ID" = SAMPLENAME,
      "Activity ID User Supplied (PARENTs)" = NA,
      "Activity Type" = "Sample-Routine",
      "Activity Media Name" = MATRIX,
      "Activity Start Date" = format(mdy_hms(SAMPDATE), "%m/%d/%Y"),
      "Activity Start Time" = format(mdy_hms(SAMPDATE), "%H:%M"),
      "Activity Start Time Zone" = "PST",
      "Activity Depth/Height Measure" = "0.152",
      "Activity Depth/Height Unit" = "m",
      "Sample Collection Method ID" = "BVR SWQAPP",
      "Sample Collection Method Context" = "CA_BVR",
      "Sample Collection Equipment Name" = "Water Bottle",
      "Sample Collection Equipment Comment" = NA,
      "Characteristic Name" = case_when(
        ANALYTE == "E. Coli" ~ "Escherichia coli",
        ANALYTE == "Oil & Grease (HEM)" ~ "Oil and Grease",
        ANALYTE == "Phosphorus, total" ~ "Phosphorus",
        ANALYTE == "Total Organic Carbon" ~ "Organic carbon",
        TRUE ~ ANALYTE
      ),
      "Characteristic Name User Supplied" = NA,
      "Method Speciation" = case_when(
        ANALYTE == "Nitrate as N" ~ "as N",
        ANALYTE == "Phosphorus, total" ~ "as P",
        ANALYTE == "Nitrate + Nitrite as N" ~ "as N",
        TRUE ~ NA_character_
      ),
      "Result Detection Condition" = case_when(
        Result == "ND" ~ "Not Reported",
        Result == "Absent" ~ "Not Present",
        Result == "Present" ~ "Detected Not Quantified",
        TRUE ~ NA_character_
      ),
      "Result Value" = ifelse(
        Result == "Absent" |
          Result == "ND" | Result == "Present",
        NA_character_,
        Result
      ),
      "Result Unit" = ifelse(UNITS == ".", NA_character_, UNITS),
      "Result Measure Qualifier" = NA,
      "Result Sample Fraction" = "Total",
      "Result Status ID" = "Final",
      "ResultTemperatureBasis" = NA,
      "Statistical Base Code" = NA,
      "ResultTimeBasis" = NA,
      "Result Value Type" = ifelse(is.na(Result), NA_character_, "Actual"),
      "Result Analytical Method ID" = ifelse(is.na(METHODNAME), NA_character_, method_id_lookup[METHODNAME]),
      "Activity ID (CHILD-subset)" = make_activity_id(
        location_id = `Monitoring Location ID`,
        date = `Activity Start Date`,
        time = `Activity Start Time`,
        activity_type = `Activity Type`,
        equipment_name = `Sample Collection Equipment Name`,
        depth = `Activity Depth/Height Measure`
      ),
      "Result Analytical Method Context" = method_context_lookup[METHODNAME],
      "Analysis Start Date" = format(mdy_hms(ANADATE), "%m/%d/%Y"),
      "Result Detection/Quantitation Limit Type" = ifelse(DL == "NA", NA_character_, "Method Detection Level"),
      "Result Detection/Quantitation Limit Measure" = ifelse(DL == "NA", NA_character_, DL),
      "Result Detection/Quantitation Limit Unit" = ifelse(UNITS == ".", NA_character_, UNITS),
      "Result Comment" = NA
      
    ) %>%
    select(-c(0:48)) %>%
    relocate("Activity ID (CHILD-subset)", .before = "Activity ID User Supplied (PARENTs)")
})

```

One last step that we need to do before getting a WQX ready dataset is to make sure that both `Result Unit` and `Result Analytical Method Context` column is empty if the `Result Analytical Method ID` and `Result Unit` columns are respectively empty. We also have to make sure that the columns are in the right order (using the function relocate() to move the columns).

The rendered table shows the WQX ready dataframe.

```{r}
clean_alpha_lab <- reactive({
  req(input$alpha_files)
  
  partial_clean_alpha_lab <-  partial_clean_alpha_lab() |>
    mutate(
      "Result Analytical Method Context" = ifelse(
        is.na(partial_clean_alpha_lab()$`Result Analytical Method ID`),
        NA_character_,
        partial_clean_alpha_lab()$`Result Analytical Method Context`
      ),
      "Result Unit" = ifelse(
        is.na(partial_clean_alpha_lab()$`Result Value`),
        NA_character_,
        partial_clean_alpha_lab()$`Result Unit`
      )
    ) |>
    relocate("Result Analytical Method Context", .before = "Analysis Start Date")
})


renderTable({
  head(clean_alpha_lab())
})

```

Lastly we export the data to a CSV file in the data folder. Cells with NAs are saved as blank cells. 

Click on the Download button for the WQX ready CSV file. 

```{r}
downloadHandler(
  filename = function() {
    paste("alpha-lab-data-", Sys.Date(), ".csv", sep = "")
  },
  content = function(file) {
    write.csv(clean_alpha_lab(),
              file,
              na = '',
              row.names = FALSE)
  }
)
```




